---
title: 'Processing raw mobility data with the `checkin` package'
tags:
  - R
  - human mobility
  - data processing
authors:
  - name: Michael Kane
    orcid: 0000-0003-1899-6662
    equal-contrib: false
    affiliation: 1 # (Multiple affiliations must be quoted)
    corresponding: true # (This is how to denote the corresponding author)
  - name: Owais Gilani
    orcid: 0000-0002-0402-6975 
    equal-contrib: false # (This is how you can denote equal contributions between multiple authors)
    affiliation: 2
  - name: Simon Urbanek
    orcid: 0000-0003-2297-1732 
    affiliation: 3
affiliations:
 - name: Biostatistics Department, School of Public Health, Yale University, New Haven, CT, USA
   index: 1
 - name: Math Department, Bucknell University, Lewisburg, PA, USA
   index: 2
 - name: Statistics Department, University of Auckland, Auckland, NZ
   index: 3
date: 26 July 2020
bibliography: paper.bib
output: md_document
---


```{r, include = FALSE}
library(future)
library(doFuture)
plan(multisession, workers = max(c(1, parallel::detectCores()-2)))
```

# Summary

The analysis of mobility data often begins with the processing of data capturing the time and location of individuals. When preparing these data for analysis, the continuous timestamp data can be particularly difficult to normalize since they require the data scientist to integrate mode assumptions such as, where is a person/device between checkins? Where is a person/device before the first checkin? Should last-observation-carried-forward be used between checkins or between descretization boundaries? For example, we may want to build sequences of checkins for a given device over locations and then count the total number of transitions between locations over all individuals.  Problems like these are common in mobility research, they require careful consideration based on the goals of an analysis, and software tools implementing these types of computations will provide benefits in terms of time savings and data integrity. To address these challenges we provide the `checkin` package, which provides a standard set of functions for appropriately descretizing spatio-timestamp data for aggregate analysis for the R programming environment [@rcore2022].


# Statement of Need

Raw mobility data is often characterized by having a column denoting the device/person identifier, a timestamp, the location, and potentially other features of the checkin. The device/person identifer is often given as a unique identifier of a device or person in possession of a device; the timestamp denotes the time at which a person was at a given location; and the location can be precise location information (such as GPS) but is often aggregated to a discrete location, such as a store, census tract, county etc. While these data are information-rich, to analyze them, especially at the aggregate level (many devices and many locations), requires processing to transform them into a representation amenable to analysis. One such representation is the _mobility graph_, which encodes vertices as discrete locations, directed edges as the aggregate movement between locations, and edge weights capturing the amount of movement (or similar measure) between locations with corresponding directed edge `[@gilani2020]`. More generally, processing steps for these aggregate analyses either descretize continuous timestamp and spatial data and/or aggregate already descretized data. 

Basic operations for processing time-stamp are implemented in the core of R and there are a plethora of package that make operations such as reading, comparing, and adding offsets more convenient (see @edd2020, @grolemund2011, etc. for more) there has not been a set of standard functions specifically for processing checking data, which requires the following types of operations:

1. Find the location of person/device at a specified interval, using last-observation-carry forward if specified.
2. Construct a generator for creating sequences of intervals over which data should be processed.
3. A map operation working in conjunction with interval iterators for processing data over spatio-interval data.

The `checkin` package provides an extensible library for providing these operations in a way that is compatible with the "tidy data" approach described in @wickham2014, it is compatible with standard `dplyr` [@wickham2022] functions, and it uses `foreach` package [@weston2022] to provide a parallel-backend for compute-intensive computing. Data sets returned from `checkin` functions are in the `tibble` [@grolemund2011] format.


# Usage

Consider the `checkins` data set from the `checkin` library shown below. The data consists of 3 columns corresponding to the device (id), time, (timestamp), and location identifier (location). There are a total of 
1000 unique people/devices rangin in time from 2020-04-19 00:01:13 EST to 2020-05-08 18:58:29 EST.

```{r}
library(checkin)
library(dplyr)
data(checkins)
checkins
```

Now suppose we would like to examine the checkins of indvidual/device number 335. We will extract all rows with id 335 and then specify an interval starting at the beginning of the hour of the first checkin to one hour later using the `checkins_in_interval()` function. The code and output are below and there are two things to note. First, the first row location is `NA`. This is because the individual/device does not appear before 2020-04-19 00:41:46 and a location cannot be determined. Second the original data set did not include an entry for id 335 at 2020-04-19 00:59:59. This value was carried forward from the previous location.

```{r}
library(lubridate)
x <- checkins %>%
  filter(id == 335) %>%
  arrange(timestamp)

start <- x$timestamp[1]
minute(start) <- 0
second(start) <- 0

end <- start + hours(1) - seconds(1)
checkins_in_interval(x, "timestamp", start, end)
```
Finally, suppose we would like to get the locations of individuals/devices at the beginning of a time interval, the location at the end of the interval, and the total amount of time the individual/devices was checked into the beginning location. This operation would be performed in two steps. First, a function, `from_to()` is constructed, which takes the rows corresponding to a single `id` for a given interval. This function finds, the initial location (`from`), the end location(`to`), the timestamp at the _beginning_ of the interval, and the duration of the initial location. In the second step, we group the `checkin` data by `id` and pass the result to the `map_hourly_interval_dfr()` function, which applies to `from_to()` function to each `id` over each hourly interval. Other intervals are included in the package and the documentation includes a information on how to construct similar functions over custom intervals.

```{r}
from_to <- function(it) {
  it$duration <- c(diff(it$timestamp), 0)
  units(it$duration) <- "secs"
  it$duration <- as.numeric(it$duration)
  from_duration <- sum(it$duration[it$location == it$location[1]])
  tibble(from=it$location[1],
         to = it$location[nrow(it)],
         timestamp = it$timestamp[1],
         from_duration = from_duration)
}

 checkins |>
   group_by(id) |>
   map_hourly_interval_dfr(from_to, time = "timestamp") 
```

# Acknowledgements

This work was partially supported by the National Science Foundation (NSF) Grant Human Networks and Data Science - Infrastructure (HNDS-I), award numbers 2024335 and 2024233.

# References
